Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-07 drafts/hybrid_zone.tex
lit review notes
peter added paragraph

Much of the process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of ancestry across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed in][]{Barton1985}. 
%\plr{(check that neutral clines discussed in barton '85)}
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
patterns of correlated ancestry in admixed populations has been previously described the history of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,sedghifar2015spatial}.

@@

\alisa{Better summary of what Barton did.
Population genetic modeling of hybrid zones has focused largely on the 
equilibrium states of clines at selected loci \citep{Nagylaki1975,Barton1979a,Barton1986,Slatkin1973,Slatkin1975,May1975,Haldane1948}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
In particular \citet{Barton1979,Barton1986} have described patterns of clinal variation at loci linked to targets of selection.  
This provides some expectation for how selection shapes covariance in ancestry at a single locus over geographic space.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry across the genome.}

>>>>

Much of the process of speciation commonly involves populations diverging with relatively little gene flow \citep{Coyne2004}.  
% Gene flow between formerly isolated populations is possible if reproductive isolation is not complete.
However, when formerly isolated populations come into contact before reproductive isolation is complete, some gene flow is possible.
%As two such populations interbreed , each portion of each individual's genome 
%can be labeled according to which ancestral population is is inherited from, and local 
Interbreeding and migration between such populations  
% YB It seems like the fact that such ancestry can be labeled is more important later, but probably shouldn't be a focus here. 
creates a gradient of ancestry across geographic space, %YB removed `assignments' as it seems like it is the ancestry, not its assignment, that is of interest
centered on their point of contact \citep[reviewed in][]{Barton1985}. 
If the populations are sufficiently diverged, 
this process leaves a distinct pattern of variation across the genome, 
in which long tracts of divergent haplotypes from each ancestral population  
are broken up by historical recombination events,
forming chromosomal \emph{junctions} in ancestry \citep{Fisher1954, Chapman2002, baird2003distribution}. 
Limited opportunities for recombination in recent hybrid zones generates substantial correlations in ancestry (and hence, genotype) between distant markers, and
patterns of correlated ancestry in admixed populations has been previously described the history of hybridization and admixture \citep[e.g.][]{Gravel2012,Hellenthal2014,sedghifar2015spatial}.

@@

There is substantial theoretical work on genetic clines.
We study those maintained \emph{only} by selection against hybrids -- \emph{tension zones} --
as in \citet{Bazykin1969}, reviewed in \citet{Barton1985}.
Clines may also be maintained by environmental gradients \citep{Haldane1948,Fisher1950};
although these are in principle different,
\citet{Slatkin1973} pointed out (among many other things)
that while the environment may determine the location of a cline,
the fitness of the heterozygote will likely determine the shape of the cline,
and \citet{Barton1986} show that the shapes of several sorts of stable clines are similar.
There is substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,Barton2008,Polechova2011},
and quantifying the barrier to gene flow caused by clines \citep{barton1979gene,barton1989adaptation,Barton1993,Barton2000}.
In this work, we focus on \emph{transient} patterns after the initial formation of a tension zone,
in particular describing haplotype patterns.
Although the selected cline establishes quickly, transient patterns can persist for quite some time,
and haplotypes can potentially carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing properties of the \emph{lineages} tracing back from the zone.
This is particularly suited to understanding transient patterns,
although because of the well-known problems with spatial coalescence \citep{felsenstein1978pain,barton2002spatial},
incorporating genetic drift is not straightforward.
Common to nearly all studies is the use of the
reaction-diffusion equations that govern the deterministic, high-density limit \citep[][as in]{Nagylaki1975},
or the discrete analogues \citep{Hansen1966}.

% \alisa{
% Population genetic modeling of hybrid zones has focused largely on the 
% equilibrium states of clines at selected loci \citep{Nagylaki1975,Barton1979a,Barton1986,Slatkin1973,Slatkin1975,May1975,Haldane1948}.  
% Since stronger selection reduces the amount of genetic mixing,
% these models predict the narrowest clines 
% (i.e., the most rapid changes in allele frequencies)
% at loci close to selected loci, and wider clines at less tightly linked loci.
% In particular \citet{Barton1979,Barton1986} have described patterns of clinal variation at loci linked to targets of selection.  
% This provides some expectation for how selection shapes covariance in ancestry at a single locus over geographic space.
% Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry across the genome.}
